---
title: "Creating a Pediatrics ADVS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Pediatrics ADVS ADaM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admiral)
library(admiraldev)
```

# Introduction

This article describes creating a vital signs ADaM for pediatric clinical trials. This package creates SD scores (z-scores) and percentiles among children and adolescents for various body measures, such as Height, Weight, Body Mass Index (BMI), Weight-for-Length, and Head Circumference.  Among adults, standard cut-points can be used (e.g., a BMI \>= 30 for obesity), but because of the large changes that occur during growth and development, these measures are typically expressed relative to other children of the same sex and age. For example, the CDC classifies obesity as a BMI-for-age \>= 95th percentile (a z-score of 1.645), while the WHO cut point is a weight-for-length \>= 3 SDs among children under 2 years of age.

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around creating the general set-up of an ADVS using
`{admiral}` functions is the same. We focus on only the pediatric-specific steps here to avoid repeating
information and maintaining the same content in two places.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input
unless otherwise specified.*

# Programming Workflow

-   [Metadata](#metadata)
-   [Initial ADVS Set-up](#advs_start)
-   [Derive Additional Variables for Anthropometric
    indicators](#derive_vars)
-   [Derive Additional Parameters for Anthropometric
    indicators](#derive_params)
-   [Remaining ADVS Set-up](#advs_end)

## Metadata {#metadata}

Once the required packages have been loaded, the first step is preparing the metadata.

```{r message=FALSE}
library(admiral)
library(admiralpeds)
library(dplyr, warn.conflicts = FALSE)
library(lubridate)
library(stringr)
```

Here, we have made some default decisions regarding which metadata to use in the
package. But you are free to replace these with any other source you prefer
(as long as you keep the structure of the dataframe consistent - as expected by our
downstream functions). For example, you might want to use a different age range selection of WHO metadata
(such as including WHO reference data for between ages [5-19
years](https://www.who.int/tools/growth-reference-data-for-5to19-years/indicators))
or the International Obesity Task Force reference data.

The selection of growth reference files we included and used as metadata
for this package are as follows:

-   [2000 CDC Growth
    Charts](https://www.cdc.gov/growthcharts/percentile_data_files.htm)
-   For BMI only: [2022 CDC Extended BMI-for-age Growth
    Charts](https://www.cdc.gov/growthcharts/extended-bmi.htm)
-   [2006 WHO Growth
    Standards](https://www.who.int/toolkits/child-growth-standards/standards) -
    Expanded Tables: Birth to 5 years

### Reference Files for Parameters by Age

As growth rates and patterns by age and sex, reference values differ by these
two characteristics. 

For the various measures, we use the WHO reference data for children <2 years 
of age (\<730.5 days) and CDC growth charts for children >=2 years of age 
(\>=730.5 days).
This is in accord with the [CDC
recommendation](https://www.cdc.gov/growthcharts/who_charts.htm). The reference 
data for both charts include the LMS parameters, in which L is the Box-Cox 
transformation for normality, M is the median, and S is the coefficient of 
variation (S). 

So, the first step is combining the metadata for each measure and ensuring sex 
and age are shown consistently.
We separate WHO and CDC metadata for weight-based derivations due to
(1) the WHO adjustment (restricted application of the LMS method) and
(2) the CDC extended method for high BMIs.

Additionally, as the CDC growth charts only offer monthly age intervals, on the
advice of a retired CDC expert in this area, we added interpolation code in our
template to have a record for each day of age.
When converting the CDC months into days (30.4375 days in a month), we rounded 
to the nearest whole number of days. Relying on the monthly charts will result 
in less accurate calculations, particularly for weight and height values among 
children who are 2 years of age. 
If you would prefer to rely solely on the monthly chart here, then you could 
remove the rounding and the call to the `derive_interp_records()` function.

We do this pre-processing of the metadata using the following code for BMI
as an example:

```{r eval=TRUE}
data(WHO_bmi_for_age_boys)
data(WHO_bmi_for_age_girls)
data(cdc_bmiage)

who_bmi_for_age <- who_bmi_for_age_boys %>%
  mutate(SEX = "M") %>%
  bind_rows(who_bmi_for_age_girls %>%
    mutate(SEX = "F")) %>%
  # Keep patients < 2 yrs old
  filter(Day < 730.5) %>%
  rename(AGE = Day) %>%
  # AGEU is added in metadata, required for derive_params_growth_age()
  mutate(AGEU = "DAYS") %>%
  arrange(AGE, SEX)

cdc_bmi_for_age <- cdc_bmiage %>%
  mutate(
    SEX = case_when(
      SEX == 1 ~ "M",
      SEX == 2 ~ "F",
      TRUE ~ NA_character_
    ),
    # Ensure first that Age unit is "DAYS"
    AGE = round(AGE * 30.4375),
    AGEU = "DAYS"
  ) %>%
  # Interpolate the AGE by SEX so that we get CDC metadata by day instead of
  # month in the same way as WHO metadata
  derive_interp_records(
    by_vars = exprs(SEX),
    parameter = "BMI"
  ) %>%
  # Keep patients >= 2 yrs till 20 yrs - Remove duplicates for 730 Days old which
  # must come from WHO metadata only
  filter(AGE >= 730.5 & AGE <= 7305) %>%
  arrange(AGE, SEX)
```

Here is how the first records of the WHO metadata for BMI now look:

```{r, eval=TRUE, echo=FALSE}
who_bmi_for_age <- who_bmi_for_age %>%
  select(AGE, AGEU, SEX, L, M, S)

dataset_vignette(
  who_bmi_for_age,
  filter = AGE < 20
)
```

Similarly, for the CDC metadata:

```{r, eval=TRUE, echo=FALSE}
cdc_bmi_for_age <- cdc_bmi_for_age %>%
  select(AGE, AGEU, SEX, L, M, S, P95, Sigma)

dataset_vignette(
  cdc_bmi_for_age,
  filter = AGE < 750
)
```

For BMI in the CDC metadata, a dispersion parameter (Sigma) is used to calculate
BMI percentiles and z-scores above the sex- and age-specific 95th percentile (P95).  

The above example only shows the parameter BMI, but this follows other parameters
like weight and height. We can combine WHO and CDC metadata for height as a
WHO adjustment is unnecessary.
Note that the head circumference parameter metadata comes only from the WHO reference data,
as CDC provides no equivalent for children \> 2 years of age.

### Reference Files for Weight by Length/Height

WHO provides additional reference data for weight-for-length (recumbent length),
instead of age. Again, we combined the metadata for sex to create a single 
reference file. 

There are also weight-by-height files available, but upon advice from a retired 
CDC expert in this area, we chose only to include the weight-for-length files, 
as once children can stand, BMI by age is a more appropriate measure to use.
However, in our template, we give the metadata values variable a generic name,
`HEIGHT_LENGTH`, so that any user could equally choose to pass weight-for-height
files if preferred. 

We do this using the following code:

```{r eval=TRUE}
data(who_wt_for_lgth_boys)
data(who_wt_for_lgth_girls)

who_wt_for_lgth <- who_wt_for_lgth_boys %>%
  mutate(SEX = "M") %>%
  bind_rows(who_wt_for_lgth_girls %>%
    mutate(SEX = "F")) %>%
  mutate(HEIGHT_LENGTHU = "cm") %>%
  rename(HEIGHT_LENGTH = Length)
```

Here is how the metadata now looks for a body length of 65cm:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  who_wt_for_lgth,
  filter = HEIGHT_LENGTH == 65
)
```

## Initial ADVS Set-up {#advs_start}

The following steps are to read in the source data, merge ADSL variables, 
and derive the usual ADVS analysis variable. The
`{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
gives detailed guidance on all these steps.

The only difference here would be the parameter level values, which now
include the pediatrics-specific parameters as needed - for example:

| `PARAMCD` | `PARAM`                | `PARAMN` | `PARCAT1`              | `PARCAT1N` |
|---------------|---------------|---------------|---------------|---------------|
| BMISDS    | BMI-for-age z-score    | 9        | Subject Characteristic | 1          |
| BMIPCTL   | BMI-for-age percentile | 10       | Subject Characteristic | 1          |

```{r echo=FALSE, message=FALSE}
data("vs_peds")
data("adsl_peds")

vs <- convert_blanks_to_na(vs_peds)
adsl <- adsl_peds %>% select(-DOMAIN)

vs <- filter(vs, USUBJID %in% c("01-701-1023"))

param_lookup <- tibble::tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM, ~PARAMN,
  "WEIGHT", "WEIGHT", "Weight (kg)", 1,
  "HEIGHT", "HEIGHT", "Height (cm)", 2,
  "BMI", "BMI", "Body Mass Index(kg/m^2)", 3,
  "HDCIRC", "HDCIRC", "Head Circumference (cm)", 4,
  NA_character_, "WGTASDS", "Weight-for-age z-score", 5,
  NA_character_, "WGTAPCTL", "Weight-for-age percentile", 6,
  NA_character_, "HGTSDS", "Height-for-age z-score", 7,
  NA_character_, "HGTPCTL", "Height-for-age percentile", 8,
  NA_character_, "BMISDS", "BMI-for-age z-score", 9,
  NA_character_, "BMIPCTL", "BMI-for-age percentile", 10,
  NA_character_, "HDCSDS", "Head Circumference-for-age z-score", 11,
  NA_character_, "HDCPCTL", "Head Circumference-for-age percentile", 12,
  NA_character_, "WGTHSDS", "Weight-for-length/height Z-Score", 13,
  NA_character_, "WGTHPCTL", "Weight-for-length/height Percentile", 14
)
attr(param_lookup$VSTESTCD, "label") <- "Vital Signs Test Short Name"

adsl_vars <- exprs(SEX, BRTHDTC, TRTSDT, TRTEDT, TRT01A, TRT01P)

advs <- vs %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = get_admiral_option("subject_keys")
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "BRTH",
    dtc = BRTHDTC
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = VSDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))

advs <- advs %>%
  derive_vars_merged_lookup(
    dataset_add = param_lookup %>% filter(!is.na(VSTESTCD)),
    new_vars = exprs(PARAMCD),
    by_vars = exprs(VSTESTCD)
  ) %>%
  mutate(AVAL = VSSTRESN)

advs <- advs %>%
  mutate(
    ATPTN = VSTPTNUM,
    ATPT = VSTPT,
    AVISIT = case_when(
      str_detect(VISIT, "UNSCHED|RETRIEVAL|AMBUL") ~ NA_character_,
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = as.numeric(case_when(
      VISIT == "SCREENING 1" ~ "-1",
      VISIT == "BASELINE" ~ "0",
      str_detect(VISIT, "WEEK") ~ str_trim(str_replace(VISIT, "WEEK", "")),
      TRUE ~ NA_character_
    ))
  )
```

## Derive Additional Variables for Anthropometric indicators {#derive_vars}

To compare against the reference files, we need to know each child's age and
length/height at the time of each vital signs assessment. 
So, these need to be first derived as variables in your ADVS.

Remember to ensure that the unit of these matches that of the reference
files metadata you are comparing against, i.e. "days" for age and "cm"
for length/height in our examples here.

Note that some growth reference files we use here have different units and intervals. 
For example when you look at the "by Age" charts, the WHO provides this by day,
but the CDC gives this every month (25.5, 26.5, etc.). So, converting all ages 
to days leaves the challenge of joining when the patient's current age falls 
between the monthly CDC reference data. In our package functions, we used the
nearest age from the metadata rather than applying some form of interpolation.
An alternative method would be to derive the current age according to the same
intervals as the metadata, with 30.4375 days in a month, for an exact join.
But, in our examples, we use days. 

### Derived Variables for Parameters by Age

A variable for current analysis age could be achieved using the
following code (you might first need to derive `BRTHDT` if not already
available in ADSL, including any partial date imputation rules if those are
possible in your study data collection rules):

```{r eval=TRUE}
# Calculate Current Analysis Age AAGECUR and unit AAGECURU
advs <- advs %>%
  derive_vars_duration(
    new_var = AAGECUR,
    new_var_unit = AAGECURU,
    start_date = BRTHDT,
    end_date = ADT
  )
```

Here is how these age variables look:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  advs,
  display_vars = exprs(USUBJID, BRTHDT, ADT, AAGECUR, AAGECURU)
)
```

### Derived Variables for Weight by Length/Height

Similar to the above, a variable for current analysis length/height
could be achieved using the following code:

```{r eval=TRUE}
# Derive Current HEIGHT/LENGTH at each time point Temporary variable
advs <- advs %>%
  derive_vars_merged(
    dataset_add = advs,
    by_vars = c(get_admiral_option("subject_keys"), exprs(AVISIT)),
    filter_add = PARAMCD == "HEIGHT" & toupper(VSSTRESU) == "CM",
    new_vars = exprs(HGTTMP = AVAL, HGTTMPU = VSSTRESU)
  )
```

## Derive Additional Parameters for Anthropometric indicators {#derive_params}

Now, we get to the most important section, which shows how to create new records
for each derived pediatrics parameter.
Only specific examples are included here, but you'll find more parameters in the
example template script referenced at the bottom of this vignette. 

### Derived Parameters by Age

Parameters for BMI-for-age z-scores and percentiles could be achieved using
the `derive_params_growth_age()` function as follows:

```{r eval=TRUE}
## Derive Anthropometric indicators (Z-Scores/Percentiles-for-Age) based on Standard Growth Charts
## For BMI by Age
advs <- advs %>%
  slice_derivation(
    derivation = derive_params_growth_age,
    args = params(
      by_vars = c(get_admiral_option("subject_keys"), exprs(AVISIT)),
      sex = SEX,
      age = AAGECUR,
      age_unit = AAGECURU,
      parameter = VSTESTCD == "BMI",
      analysis_var = AVAL,
      set_values_to_sds = exprs(
        PARAMCD = "BMISDS",
        PARAM = "BMI-for-age z-score"
      ),
      set_values_to_pctl = exprs(
        PARAMCD = "BMIPCTL",
        PARAM = "BMI-for-age percentile"
      )
    ),
    derivation_slice(
      filter = AAGECUR < 730.5,
      args = params(
        who_correction = TRUE,
        meta_criteria = who_bmi_for_age
      )
    ),
    derivation_slice(
      filter = AAGECUR >= 730.5,
      args = params(
        bmi_cdc_correction = TRUE,
        meta_criteria = cdc_bmi_for_age
      )
    )
  )
```

If only a z-score or percentile were needed, you leave out one of the `set_values_to_`
arguments.

Once again, it is essential to remember the importance of the metadata being
supplied to the above function via the `meta_criteria` argument.
Our default case from this package uses WHO standards for children \< 2 years of
age and CDC reference data for children \>= 2 years of age. As the user, you can 
use other (e.g., IOTF) reference data files created with the LMS method.

For BMI only, you'll notice from the code above that we have
set `bmi_cdc_correction = TRUE`. This is because we used the CDC
extended percentiles (\> 95th percentile) to monitor high BMI values here.
If you left this argument as default `FALSE` value, only the 2000 CDC Growth 
Chart for BMI would be used from the CDC metadata. In 2022, the CDC released 
extended percentiles (and z-scores) for BMIs \> 95th percentile because the 
original CDC method did not accurately characterize very high BMIs. 

For all weight-based parameters (including BMI), you'll also see in the code
above that we have set `who_correction = TRUE`. This comes from the
[WHO Child Growth Standards Guidelines](https://www.who.int/publications/i/item/924154693X)
(pages 301-304). This correction is made because the WHO used the LMS method only
for weight-based measures for which the calculated z-scores were
between -3 and +3; this correction allows for the calculation of more
extreme z-scores. If you are not using the WHO metadata 
(or WHO metadata for length or head circumference), you could leave
`who_correction = FALSE`, and only the usual LMS method would be applied.
WHO recommends using this correction for all parameters that involve weight (including BMI).  

Here is how these newly derived parameters look:

```{r, eval=TRUE, echo=FALSE}
advs_display <- advs %>% select(USUBJID, AAGECUR, AAGECURU, PARAMCD, PARAM, AVAL)

dataset_vignette(
  advs_display,
  filter = PARAMCD %in% c("BMISDS", "BMIPCTL")
)
```

For height parameters by age, it should be noted that the WHO growth chart
we use for children \<2 years of age refers to body length, whereas
the CDC growth chart for children \>=2 years of age refers to height.
This is explained thoroughly in the next section.

### Derived Weight by Length/Height

The following code is for Weight by Length/Height z-score and percentile for
children \< 2 years of age, which assumes the "HEIGHT" parameter is always
collected as recumbent length. 

You'll notice the `derive_params_growth_height()` function used is very similar
to that used above, but now passing in height or length instead of age.

```{r eval=FALSE}
advs <- advs %>%
  restrict_derivation(
    derivation = derive_params_growth_height,
    args = params(
      by_vars = c(get_admiral_option("subject_keys"), exprs(AVISIT)),
      sex = SEX,
      height = HGTTMP,
      height_unit = HGTTMPU,
      meta_criteria = who_wt_for_lgth,
      parameter = VSTESTCD == "WEIGHT",
      analysis_var = AVAL,
      who_correction = TRUE,
      set_values_to_sds = exprs(
        PARAMCD = "WGTHSDS",
        PARAM = "Weight-for-length/height Z-Score"
      ),
      set_values_to_pctl = exprs(
        PARAMCD = "WGTHPCTL",
        PARAM = "Weight-for-length/height Percentile"
      )
    ),
    filter = AAGECUR < 730.5
  )
```

We consulted with a pediatrician with extensive experience working with length 
and height data, who advised that analysts rarely know whether standing height 
or recumbent length was measured. It depends mainly on the age at which the 
child can stand (on average, two years) and the preference of the measurer.

If you have strict CRF (case report form) guidelines that standing height
(rather than body length) was measured, you could add 0.7 cm to these values to
approximate the child’s body length.

## Remaining ADVS Set-up {#advs_end}

The `{admiral}`
[Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html) 
covers all remaining steps, such as merging the parameter-level values, 
variables, and analysis flags.

# Example Scripts {#example}

| ADaM | Sourcing Command                                            |
|------|-------------------------------------------------------------|
| ADVS | `admiral::use_ad_template("ADVS", package = "admiralpeds")` |
