---
title: "Creating a Pediatrics ADVS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Pediatrics ADVS ADaM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admiral)
library(admiraldev)
```

# Introduction

This article describes creating a vital signs ADaM for pediatric
clinical trials. The main focus is around Anthropometric indicators
(i.e. growth development chart) parameter derivations for Height, Weight, Body
Mass Index (BMI), and Head Circumference.

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around how to create the general set-up of an
ADVS using `{admiral}` functions is the same as this, so to save
repeating information and maintaining the same content in two places we
have instead opted to only focus on the pediatric-specific steps here.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input
unless otherwise specified.*

# Programming Workflow

-   [Metadata](#metadata)
-   [Initial ADVS Set-up](#advs_start)
-   [Derive Additional Variables for Anthropometric
    indicators](#derive_vars)
-   [Derive Additional Parameters for Anthropometric
    indicators](#derive_params)
-   [Remaining ADVS Set-up](#advs_end)

## Metadata {#metadata}

Once the required packages have been loaded, the first step is then to
prepare the metadata.

```{r message=FALSE}
library(admiral)
library(admiralpeds)
library(dplyr, warn.conflicts = FALSE)
library(lubridate)
library(rlang)
library(stringr)
```

Here we have made some default decisions in the package as to which
metadata to use but you are free to replace these with any other source
you prefer (as long as you keep the structure of the dataframe
consistent - as expected by our downstream functions). For example, you
might want to use a different age range selection of WHO metadata
(such as including WHO reference data for between ages [5-19
years](https://www.who.int/tools/growth-reference-data-for-5to19-years/indicators))
or even alternatively the International Obesity Task Force standards.

The selection of standard reference files we included and used as
metadata for this package are as follows:

-   [2000 CDC Growth
    Charts](https://www.cdc.gov/growthcharts/percentile_data_files.htm)
-   For BMI only: [2022 CDC Extended BMI-for-age Growth
    Charts](https://www.cdc.gov/growthcharts/extended-bmi.htm)
-   [2006 WHO Growth
    Standards](https://www.who.int/toolkits/child-growth-standards/standards) -
    Expanded Tables: Birth to 5 years

### Reference Files for Parameters by Age

As growth rates and patterns differ between sex, different reference
values are used accordingly and for different age groups.

For the applicable parameters here (height/weight/BMI), we made the
default decision to rely on the WHO standards for children \<2 years of
age (\<730.5 days) and CDC standards for children \>=2 years of age
(\>=730.5 days). This is made in accordance with the [CDC
recommendation](https://www.cdc.gov/growthcharts/who_charts.htm).

So, the first step here needed is to combine the various metadata by
parameter and ensuring sex and age are shown in a consistent way.
For weight-based indicators we keep WHO and CDC metadata separate due to the WHO
adjustment (restricted application of the LMS method) needed for the
anthropometric indicator derivations.

Additionally, as the CDC growth charts only offer monthly intervals we
took advice from an ex-CDC expert in this area and made the decision in
our template to add interpolation code so that we have a record for each
day of age. When converting the CDC months into days we use rounding
to bring them to a whole number of days. If you would prefer to rely
solely on the monthly chart here, then you could remove the rounding
and the call to the `derive_interp_records()` function.

We do this pre-processing of the metadata using the following code for BMI
as an example:

```{r eval=TRUE}
data(WHO_bmi_for_age_boys)
data(WHO_bmi_for_age_girls)
data(cdc_bmiage)

who_bmi_for_age <- who_bmi_for_age_boys %>%
  mutate(SEX = "M") %>%
  bind_rows(who_bmi_for_age_girls %>%
    mutate(SEX = "F")) %>%
  # Keep patients < 2 yrs old
  filter(Day < 730.5) %>%
  rename(AGE = Day) %>%
  # AGEU is added in metadata, required for derive_params_growth_age()
  mutate(AGEU = "DAYS") %>%
  arrange(AGE, SEX)

cdc_bmi_for_age <- cdc_bmiage %>%
  mutate(
    SEX = case_when(
      SEX == 1 ~ "M",
      SEX == 2 ~ "F",
      TRUE ~ NA_character_
    ),
    # Ensure first that Age unit is "DAYS"
    AGE = round(AGE * 30.4375),
    AGEU = "DAYS"
  ) %>%
  # Interpolate the AGE by SEX so that we get CDC metadata by day instead of
  # month in the same way as WHO metadata
  derive_interp_records(
    by_vars = exprs(SEX),
    parameter = "BMI"
  ) %>%
  # Keep patients >= 2 yrs till 20 yrs - Remove duplicates for 730 Days old which
  # must come from WHO metadata only
  filter(AGE >= 730.5 & AGE <= 7305) %>%
  arrange(AGE, SEX)
```

Here is how the first records of the WHO metadata for BMI now look:

```{r, eval=TRUE, echo=FALSE}
who_bmi_for_age <- who_bmi_for_age %>%
  select(AGE, AGEU, SEX, L, M, S)

dataset_vignette(
  who_bmi_for_age,
  filter = AGE < 20
)
```

As well as similar for the CDC metadata:

```{r, eval=TRUE, echo=FALSE}
cdc_bmi_for_age <- cdc_bmi_for_age %>%
  select(AGE, AGEU, SEX, L, M, S, P95, Sigma)

dataset_vignette(
  cdc_bmi_for_age,
  filter = AGE < 750
)
```

To help interpret this metadata: the LMS parameters are the median (M), the
generalized coefficient of variation (S), and the power in the Box-Cox
transformation (L).

For CDC metadata only, Sigma is the dispersion parameter used in the calculation
of BMI percentiles and z-scores above the 95th percentile, and P95 represents
this sex and age-specific 95th percentile.

The above example only shows the parameter BMI, but similar follows for the
other parameters like weight and height (except no Sigma and P95 for those),
and for height we are able to combine WHO and CDC metadata as there is no WHO
adjustment needed.
Note that the head circumference parameter metadata comes only from
the WHO standards, as no equivalent for \>2 years of age are provided by CDC.

### Reference Files for Weight by Length/Height

WHO provides additional standards for weight by body length (recumbent length),
instead of age. For these we again combined the metadata for sex so that we have
a single reference file. 

There are also weight by height files available, but upon advice from an ex-CDC
expert in this area we chose to only include the by body length files, as once
children are able to stand then BMI by age is a more appropriate measure to use.
However, in our template we do give the metadata values variable a more generic
name as `HEIGHT_LENGTH` so that any user could equally choose to pass weight by
height files if they so preferred.

We do this using the following code:

```{r eval=TRUE}
data(who_wt_for_lgth_boys)
data(who_wt_for_lgth_girls)

who_wt_for_lgth <- who_wt_for_lgth_boys %>%
  mutate(SEX = "M") %>%
  bind_rows(who_wt_for_lgth_girls %>%
    mutate(SEX = "F")) %>%
  mutate(HEIGHT_LENGTHU = "cm") %>%
  rename(HEIGHT_LENGTH = Length)
```

Here is how the metadata now looks for a body length of 65cm:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  who_wt_for_lgth,
  filter = HEIGHT_LENGTH == 65
)
```

## Initial ADVS Set-up {#advs_start}

The next step would be reading in the source data, merging ADSL
variables and deriving the usual ADVS analysis variables. The
`{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
gives detailed guidance on all these steps.

The only difference here would be the parameter level values, which now
include the pediatrics-specific parameters as needed - for example:

| `PARAMCD` | `PARAM`                | `PARAMN` | `PARCAT1`              | `PARCAT1N` |
|---------------|---------------|---------------|---------------|---------------|
| BMISDS    | BMI-for-age z-score    | 9        | Subject Characteristic | 1          |
| BMIPCTL   | BMI-for-age percentile | 10       | Subject Characteristic | 1          |

```{r echo=FALSE, message=FALSE}
data("vs_peds")
data("adsl_peds")

vs <- convert_blanks_to_na(vs_peds)
adsl <- adsl_peds %>% select(-DOMAIN)

vs <- filter(vs, USUBJID %in% c("01-701-1023"))

param_lookup <- tibble::tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM, ~PARAMN,
  "WEIGHT", "WEIGHT", "Weight (kg)", 1,
  "HEIGHT", "HEIGHT", "Height (cm)", 2,
  "BMI", "BMI", "Body Mass Index(kg/m^2)", 3,
  "HDCIRC", "HDCIRC", "Head Circumference (cm)", 4,
  NA_character_, "WGTASDS", "Weight-for-age z-score", 5,
  NA_character_, "WGTAPCTL", "Weight-for-age percentile", 6,
  NA_character_, "HGTSDS", "Height-for-age z-score", 7,
  NA_character_, "HGTPCTL", "Height-for-age percentile", 8,
  NA_character_, "BMISDS", "BMI-for-age z-score", 9,
  NA_character_, "BMIPCTL", "BMI-for-age percentile", 10,
  NA_character_, "HDCSDS", "Head Circumference-for-age z-score", 11,
  NA_character_, "HDCPCTL", "Head Circumference-for-age percentile", 12,
  NA_character_, "WGTHSDS", "Weight-for-length/height Z-Score", 13,
  NA_character_, "WGTHPCTL", "Weight-for-length/height Percentile", 14
)
attr(param_lookup$VSTESTCD, "label") <- "Vital Signs Test Short Name"

adsl_vars <- exprs(SEX, BRTHDTC, TRTSDT, TRTEDT, TRT01A, TRT01P)

advs <- vs %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = get_admiral_option("subject_keys")
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "BRTH",
    dtc = BRTHDTC
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = VSDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))

advs <- advs %>%
  derive_vars_merged_lookup(
    dataset_add = param_lookup %>% filter(!is.na(VSTESTCD)),
    new_vars = exprs(PARAMCD),
    by_vars = exprs(VSTESTCD)
  ) %>%
  mutate(AVAL = VSSTRESN)

advs <- advs %>%
  mutate(
    ATPTN = VSTPTNUM,
    ATPT = VSTPT,
    AVISIT = case_when(
      str_detect(VISIT, "UNSCHED|RETRIEVAL|AMBUL") ~ NA_character_,
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = as.numeric(case_when(
      VISIT == "SCREENING 1" ~ "-1",
      VISIT == "BASELINE" ~ "0",
      str_detect(VISIT, "WEEK") ~ str_trim(str_replace(VISIT, "WEEK", "")),
      TRUE ~ NA_character_
    ))
  )
```

## Derive Additional Variables for Anthropometric indicators {#derive_vars}

In order to be able to compare against the reference files we need to
know each child's age and length/height respectively at the time of each
vital signs assessment. So these need to be first derived as variables
in your ADVS.

Remember to ensure that the unit of these matches that of the reference
files metadata you are comparing against, i.e. "days" for age and "cm"
for length/height in our examples here.

Note that some of the standard reference files we use here have
different units and intervals themselves. For example when you look at
the "by Age" charts WHO provides this by day but CDC gives this every
half month. So when converting all as days this still leaves the
challenge of how to join when the patient's current age falls between an
interval (e.g. between 24.5 and 25 months for CDC). Here in our package
functions we took the simple approach to use the nearest age from the
metadata, rather than apply some form of interpolation. An alternative
method you could use would be to derive current age according to the
exact same intervals as the metadata in order for an exact join, but in
our examples we just kept this by days.

### Derived Variables for Parameters by Age

A variable for current analysis age could be achieved using the
following code (you might first need to derive `BRTHDT` if not already
available in ADSL, including any partial date imputation rules if those are
possible in your study data collection rules):

```{r eval=TRUE}
# Calculate Current Analysis Age AAGECUR and unit AAGECURU
advs <- advs %>%
  derive_vars_duration(
    new_var = AAGECUR,
    new_var_unit = AAGECURU,
    start_date = BRTHDT,
    end_date = ADT
  )
```

Here is how these age variables look:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  advs,
  display_vars = exprs(USUBJID, BRTHDT, ADT, AAGECUR, AAGECURU)
)
```

### Derived Variables for Weight by Length/Height

Similar to the above, a variable for current analysis length/height
could be achieved using the following code:

```{r eval=TRUE}
# Derive Current HEIGHT/LENGTH at each time point Temporary variable
advs <- advs %>%
  derive_vars_merged(
    dataset_add = advs,
    by_vars = c(get_admiral_option("subject_keys"), exprs(AVISIT)),
    filter_add = PARAMCD == "HEIGHT" & toupper(VSSTRESU) == "CM",
    new_vars = exprs(HGTTMP = AVAL, HGTTMPU = VSSTRESU)
  )
```

## Derive Additional Parameters for Anthropometric indicators {#derive_params}

Now we get to the most important section where we show how to create the
new records for each derived pediatrics parameter. Only certain examples
are included here, but you'll find more parameters in the example
template script referenced at the bottom of this vignette.

### Derived Parameters by Age

Parameters for BMI by Age z-score and percentile could be achieved using
the `derive_params_growth_age()` function as follows:

```{r eval=TRUE}
## Derive Anthropometric indicators (Z-Scores/Percentiles-for-Age) based on Standard Growth Charts
## For BMI by Age
advs <- advs %>%
  slice_derivation(
    derivation = derive_params_growth_age,
    args = params(
      sex = SEX,
      age = AAGECUR,
      age_unit = AAGECURU,
      parameter = VSTESTCD == "BMI",
      analysis_var = AVAL,
      set_values_to_sds = exprs(
        PARAMCD = "BMISDS",
        PARAM = "BMI-for-age z-score"
      ),
      set_values_to_pctl = exprs(
        PARAMCD = "BMIPCTL",
        PARAM = "BMI-for-age percentile"
      )
    ),
    derivation_slice(
      filter = AAGECUR < 730.5,
      args = params(
        who_correction = TRUE,
        meta_criteria = who_bmi_for_age
      )
    ),
    derivation_slice(
      filter = AAGECUR >= 730.5,
      args = params(
        bmi_cdc_correction = TRUE,
        meta_criteria = cdc_bmi_for_age
      )
    )
  )
```

In the rare case where only z-score or percentile was needed and not
both then you could leave out either of the respective `set_values_to_`
arguments.

Here it is important once again to remind of the importance of the
metadata being supplied to the above function via the `meta_criteria`
argument. In our default case from this package this is taking WHO
standards for children \<2 years of age and CDC standards for children
\>=2 years of age. You as the user are free to supply any other choice
of metadata here from the reference files you prefer.

For BMI parameters only, you'll notice from the code above that we have
set `bmi_cdc_correction = TRUE`. This is because we opted to use the CDC
developed extended percentiles (\>95%) to monitor high BMI values here.
If you left this argument as default `FALSE` value then from the CDC metadata
only the 2000 CDC Growth Chart for BMI would be used.

For any weight-based indicator (including BMI), you'll also see in the code
above that we have set `who_correction = TRUE`. This comes from the
[WHO Child Growth Standards Guidelines](https://www.who.int/publications/i/item/924154693X)
(see document page 302). The rationale is that WHO constructed a restricted
application of the LMS method here.
Of course if you were choosing to not use WHO metadata then you could
equally leave `who_correction = FALSE` and only the usual LMS method would
be applied.

Here is how these newly derived parameters look:

```{r, eval=TRUE, echo=FALSE}
advs_display <- advs %>% select(USUBJID, AAGECUR, AAGECURU, PARAMCD, PARAM, AVAL)

dataset_vignette(
  advs_display,
  filter = PARAMCD %in% c("BMISDS", "BMIPCTL")
)
```

For height parameters by age, it should be noted that the WHO growth chart
that we use for children \<2 years of age refers to body length, whereas
the CDC growth chart for children \>=2 years of age refers to height.
This is explained thoroughly in the next section.

### Derived Weight by Length/Height

Parameters for Weight by Length/Height z-score and percentile for children under
2 years of age could be achieved using the following code, where it assumes the
"HEIGHT" parameter is always collected as body length (recumbent length).

You'll notice the `derive_params_growth_height()` function used is very similar
to that used above, but now passing in height instead of age.

```{r eval=FALSE}
advs <- advs %>%
  restrict_derivation(
    derivation = derive_params_growth_height,
    args = params(
      sex = SEX,
      height = HGTTMP,
      height_unit = HGTTMPU,
      meta_criteria = who_wt_for_lgth,
      parameter = VSTESTCD == "WEIGHT",
      analysis_var = AVAL,
      who_correction = TRUE,
      set_values_to_sds = exprs(
        PARAMCD = "WGTHSDS",
        PARAM = "Weight-for-length/height Z-Score"
      ),
      set_values_to_pctl = exprs(
        PARAMCD = "WGTHPCTL",
        PARAM = "Weight-for-length/height Percentile"
      )
    ),
    filter = AAGECUR < 730.5
  )
```

We consulted with an acting pediatrician who advised that rarely investigators
are guided as to which measure to use. It depends mainly on the age
at which the child is able to stand (2 years on average) as well
as the preference of the measurer.

If you have strict CRF (case report form) guidelines that standing height
(rather than body length) was measured, you could add 0.7 cm to these values to
approximate the child’s body length. Or alternatively you could provide a
different source of reference files by height which could allow for this
parameter derivation for children \>2 years of age, but this is not our package
recommendation.

## Remaining ADVS Set-up {#advs_end}

All remaining steps such as merging on the parameter level values and
variables such as analysis flags are all covered in the `{admiral}`
[Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).

# Example Scripts {#example}

| ADaM | Sourcing Command                                            |
|------|-------------------------------------------------------------|
| ADVS | `admiral::use_ad_template("ADVS", package = "admiralpeds")` |
