---
title: "Creating a Pediatrics ADVS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Pediatrics ADVS ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating a vital signs ADaM for pediatric clinical
trials. The main focus is around growth development chart parameter
derivations.

We advise you first consult the `{admiral}`
[Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around how to create the general set-up of an
ADVS using `{admiral}` functions is the same as this, so to save repeating
information and maintaining the same content in two places we have instead
opted to only focus on the pediatric-specific steps here.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Workflow

* [Metadata](#metadata)
* [Initial ADVS Set-up](#advs_start)
* [Derive Additional Variables for Pediatrics Measures](#derive_vars)
* [Derive Additional Parameters for Pediatrics Measures](#derive_params)
* [Remaining ADVS Set-up](#advs_end)

## Metadata {#metadata}

Once the required packages have been loaded, the first step is then to prepare
the metadata.

```{r message=FALSE}
library(admiral)
library(admiralpeds)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
library(stringr)
library(tibble)
```

Here we have made some default decisions in the package as to which metadata
to use but you are free to replace these with any other source you prefer (as
long as you keep the structure of the dataframe consistent - as expected by
our downstream functions). For example, you might want to use a different
age range selection of WHO metadata or even use International Obesity Task Force
standards.

The selection of standard reference files we included and used as metadata for
this package are as follows:

- [2000 CDC Growth Charts](https://www.cdc.gov/growthcharts/percentile_data_files.htm)
- For BMI only: [2022 CDC Extended BMI-for-age Growth Charts](https://www.cdc.gov/growthcharts/extended-bmi.htm)
- [2006 WHO Growth Standards](https://www.who.int/toolkits/child-growth-standards/standards) - Expanded Tables

### Reference Files for Parameters by Age

As growth rates and patterns differ between gender, different reference values are
used accordingly and for different age groups.

For the applicable parameters here (height/weight/BMI), we made the default decision
to rely on the WHO standards for children <2 years of age (<730.5 days) and CDC
standards for children >=2 years of age (>=730.5 days). This is made in accordance
with the [CDC recommendation](https://www.cdc.gov/growthcharts/who_charts.htm).

So, the first step here needed is to combine the various metadata by parameter
and ensuring gender and age are shown in a consistent way. We do this using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE (1 example only!)
```

Here is how the combined metadata now looks for the age ranges around 2 years
to show how the WHO and CDC standards have been combined:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  bmi_for_age,
  filter = 728 < AGE < 750
)
```

Note that head circumference metadata comes only from the WHO standards, as
no equivalent for >2 years of age are provided by CDC.

### Reference Files for Weight by Length/Height

WHO provides additional standards for weight by length/height instead of
age. For these we again combined the metadata for gender and additionally
made clear which records refer to length vs height. We do this using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

Here is how the combined metadata now looks for length or height of 65cm
to show how the by length and by height standards have been combined:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  who_wt_for_ht_lgth,
  filter = HEIGHT_LENGTH == 65
)
```

## Initial ADVS Set-up {#advs_start}

The next step would be reading in the source data, merging ADSL variables
and deriving the usual  ADVS analysis variables. The `{admiral}`
[Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
gives detailed guidance on all these steps.

The only difference here would be the parameter level values, which now
include the pediatrics-specific parameters as needed - for example:

`PARAMCD` | `PARAM` | `PARAMN` | `PARCAT1` | `PARCAT1N`
--------- | -------- | ------- | --------- | ----------
BMISDS | BMI-for-age z-score | 7 | Subject Characteristic | 1
BMIPCTL | BMI-for-age percentile | 8 | Subject Characteristic | 1

# NEED TO ADD ALL THE CODE BLOCKS FROM THE TEMPLATE AND DECIDE WHETHER NECESSARY TO SHOW ANY IN THE VIGNETTE OR
# JUST RUN THEM IN THE BACKGROUND (AS MOST ARE REPEATS FROM ADMIRAL ADVS TEMPLATE AND VIGNETTE ANYWAY)

```{r eval=TRUE}
data("adsl_peds")
data("vs_peds")

adsl <- adsl_peds
vs <- convert_blanks_to_na(vs_peds)
```
```{r echo=FALSE}
vs <- filter(vs, USUBJID %in% c("01-701-1015", "01-701-1023", "01-703-1086", "01-703-1096", "01-707-1037", "01-716-1024"))
```

## Derive Additional Variables for Pediatrics Measures {#derive_vars}

In order to be able to compare against the reference files we need to
know each child's age or length/height respectively at the time of each
vital signs assessment. So these need to be first derived as variables
in your ADVS.

Remember to ensure that the unit of these matches that of the
reference files metadata you are comparing against, i.e. "days" for
age and "cm" for length/height in our examples here.

### Derived Variables for Parameters by Age

A variable for current analysis age could be achieved using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

### Derived Variables for Weight by Length/Height

A variable for current analysis length/height could be achieved using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

## Derive Additional Parameters for Pediatrics Measures {#derive_params}

Now we get to the most important section where we show how to create the
new records for each derived pediatrics parameter. Only certain examples
are included here, but you'll find more parameters in the example template
script.

### Derived Parameters by Age

Parameters for BMI by Age z-score and percentile could be achieved using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

In the rare case where only z-score or percentile was needed and not both
then you could leave out either of the respective `set_values_to_` arguments.

Here it is important once again to remind of the importance of the metadata
being supplied to the above function via the `meta_criteria` argument. In
our default case from this package this is taking WHO standards for children
<2 years of age and CDC standards for children >=2 years of age. You as the
user are free to supply any other choice of metadata here from the reference
files you prefer.

### Derived Weight by Length/Height

Parameters for Weight by Length/Height z-score and percentile could be
achieved using the following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

This function is slightly more complex as we had to add flexibility to
allow for different data collection rules around whether body length or
height was collected. Given each has it's own reference file. In our
case we consulted with an acting pediatrician who advised that rarely
they are guided as to which to take and it depends mainly on the age
at which the child is able to stand (2 years being the average) as
well as the preference of the measurer. So here we have assumed that
what was provided in the "HEIGHT" data was actually body length up to
the age of 2 and then height from then onwards. If you have strict
CRF (case report form) guidelines here then you could adjust the above
rules via the `height_age` argument. For example, if body length was
always to be taken until children are 1000 days old then you'd set
`height_age = 1000`.

However, if your pediatrics trial for example was in older patients
where true height was only ever collected (instead of body length) then
you could leave the `height_age` argument NULL and instead only supply
the by height reference files in the metadata as per the `meta_criteria`
argument.

## Remaining ADVS Set-up {#advs_end}

All remaining steps such as merging on the parameter level values and
variables such as analysis flags are all covered in the `{admiral}`
[Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).

# Example Scripts {#example}

ADaM | Sourcing Command
---- | --------------
ADVS | `admiral::use_ad_template("ADVS", package = "admiralpeds")`

Reminders:
See outline doc notes
ADD NOTE from David for BMI: CDC BMI percentiles >95 are calculated differently than this <95th percentile.
