---
title: "Creating a Pediatrics ADVS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Pediatrics ADVS ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating a vital signs ADaM for pediatric clinical
trials. The main focus is around growth development chart parameter
derivations.

We advise you first consult the `{admiral}`
[Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around how to create the general set-up of an
ADVS using `{admiral}` functions is the same as this, so to save repeating
information and maintaining the same content in two places we have instead
opted to only focus on the pediatric-specific steps here.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Workflow

* [Metadata](#metadata)
* [Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`](#paramcd)
* [Derive Additional Parameters for Pediatrics Measures](#derive_param)

## Metadata {#metadata}

Once the required packages have been loaded, the first step is then to prepare
the metadata.

```{r message=FALSE}
library(admiral)
library(admiralpeds)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
library(stringr)
library(tibble)
```

Here we have made some default decisions in the package as to which metadata
to use but you are free to replace these with any other source you prefer (as
long as you keep the structure of the dataframe consistent - as expected by
our downstream functions). For example, you might want to use a different
age range selection of WHO metadata or even use International Obesity Task Force
standards.

The selection of standard reference files we included and used as metadata for
this package are as follows:

- [2000 CDC Growth Charts](https://www.cdc.gov/growthcharts/percentile_data_files.htm)
- For BMI only: [2022 CDC Extended BMI-for-age Growth Charts](https://www.cdc.gov/growthcharts/extended-bmi.htm)
- [2006 WHO Growth Standards](https://www.who.int/toolkits/child-growth-standards/standards) - Expanded Tables

### Reference Files for Parameters by Age

As growth rates and patterns differ between gender, different reference values are
used accordingly and for different age groups.

For the applicable parameters here (height/weight/BMI), we made the default decision
to rely on the WHO standards for children <2 years of age (<730.5 days) and CDC
standards for children >=2 years of age (>=730.5 days). This is made in accordance
with the [CDC recommendation](https://www.cdc.gov/growthcharts/who_charts.htm).

So, the first step here needed is to combine the various metadata by parameter
and ensuring gender and age are shown in a consistent way. We do this using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE (1 example only!)
```

Here is how the combined metadata now looks for the age ranges around 2 years
to show how the WHO and CDC standards have been combined:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  bmi_for_age,
  filter = 728 < AGE < 750
)
```

Note that head circumference metadata comes only from the WHO standards, as
no equivalent for >2 years of age are provided by CDC.

### Reference Files for Weight by Length/Height

WHO provides additional standards for weight by length/height instead of
age. For these we again combined the metadata for gender and additionally
made clear which records refer to length vs height. We do this using the
following code:

```{r eval=TRUE}
# ADD CODE FROM TEMPLATE
```

Here is how the combined metadata now looks for length or height of 65cm
to show how the by length and by height standards have been combined:

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  who_wt_for_ht_lgth,
  filter = HEIGHT_LENGTH == 65
)
```

ADD

DATA SECTION
     
```{r eval=TRUE}
data("adsl_peds")
data("vs_peds")

adsl <- adsl_peds
vs <- convert_blanks_to_na(vs_peds)
```
```{r echo=FALSE}
vs <- filter(vs, USUBJID %in% c("01-701-1015", "01-701-1023", "01-703-1086", "01-703-1096", "01-707-1037", "01-716-1024"))
```




Reminders:
See outline doc notes
Update contents
ADD NOTE from David for BMI: CDC BMI percentiles >95 are calculated differently than this <95th percentile.
Explain any unit conversion ever needed for AGE, HEIGHT etc
Explain height and body length usual conventions and update according to your CRF guidelines
